Process-based phenotypes: an application to the ontology of cell physiology

### 2011-11-27 paper/drinking.tex

Here, we present the foundations for an ontology of process
phenotypes. We present the outline of a theory in which several kinds
of process attributes can be distinguished so that normal and abnormal
physiology of biological systems can be formally characterized. We
apply this theory to cellular processes and create the Cell Phenotype
Ontology (CPO). The CPO is linked to reference ontologies for
qualities, biological processes, functions and cell components, and
its prime application is the unification of phenotypes on the cellular
level across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists.

@@

In addition to classes and individuals, ontologies often include {\em
  relations}. Relations are the ``the glue that holds things together,
the primary constituents of the facts that go to make up reality''
\cite{}.

@@

\subsection{Cellular Phenotype Ontology}
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth,
  height=.33\textheight]{overview.pdf}
  \caption{Overview over ...\label{fig:overview}}
\end{figure*}

@@

\subsection{Cell Phenotype Ontology: implementation}
We were faced with two choices for implementing the CPO: we could
either implement a pre-composed ontology in which all classes and
their definitions are pre-generated according to the patterns we
define, or we could develop an annotation software that enables the
selection of our process phenotype patterns based on the current
structure of the GO.  To maximize the utility and compatibility of the
CPO, and to provide stable identifiers for its concepts, we selected
the first strategy and developed a software to automatically generate
a pre-composed ontology from the GO.

@@

\subsection{Pre- vs. post-composed ontology}
Pre-composed or software support for pattern?

\subsection{Formalization}
Formal, axiomatic system for process qualities, maybe based on PSL,
FOIS paper 2010

>>>>

Here, we present the foundations for an ontology of process
phenotypes. We present the outline of a theory in which several kinds
of process attributes can be distinguished so that normal and abnormal
physiology of biological systems can be formally characterized. We
apply this theory to cellular processes and create the Cell Phenotype
Ontology (CPO). The CPO is linked to reference ontologies for
qualities, biological processes, functions and cell components, and
its prime application is the unification of phenotypes on the cellular
level across different species as well as for annotation of cellular
phenotypes in domains in which no such ontology exists...

@@

In addition to classes and individuals, ontologies often include {\em
  relations}. Relations are the ``the glue that holds things together,
the primary constituents of the facts that go to make up reality''
\cite{Barwise1989}.

@@

\subsection{Cell Phenotype Ontology}
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth,
  height=.33\textheight]{overview.pdf}
  \caption{Overview over ...\label{fig:overview}}
\end{figure*}

@@

\subsection{Implementation}
We were faced with two choices for implementing the CPO: we could
either implement a pre-composed ontology in which all classes and
their definitions are pre-generated according to the patterns we
define, or we could develop an annotation software that enables the
selection of our process phenotype patterns based on the current
structure of the GO.  To maximize the utility and compatibility of the
CPO, and to provide stable identifiers for its concepts, we selected
the first strategy and developed a software to automatically generate
a pre-composed ontology from the GO.

@@

\subsection{Pre- vs. post-composed ontology}
We implemented the CPO using only a few patterns for formulating
phenotypes involving processes.

\subsection{Future work}
Formal, axiomatic system for process qualities, maybe based on PSL,
FOIS paper 2010
